# (PART) Appendix {-}

# Related software

Below you can find all the related software to `SpatialExperiment` available on Bioconductor or CRAN.

```{r SpatialExperiment_rev_dep, echo = FALSE, message = FALSE}
## Related to https://github.com/r-lib/devtools/issues/2368
packages_all <- function() {
  repos <- BiocManager::repositories()
  views <-
    paste(repos[grepl("BioC", names(repos))], "VIEWS", sep = "/")
  views <- views[!grepl("books", views)]
  
  all_pkgs <-
    c(
      lapply(views, devtools:::bioc_packages),
      list("CRAN" = devtools:::cran_packages())
    )
  
  Reduce(function(...)
    merge(..., all = TRUE), all_pkgs)
  
}

revdep_all <-
  function(pkg,
    dependencies = c("Depends", "Imports", "Suggests",
      "LinkingTo"),
    recursive = FALSE,
    ignore = NULL,
    bioconductor = FALSE)
  {
    if (missing(pkg))
      pkg <- as.package(".")$package
    all <- if (bioconductor)
      packages_all()
    else
      devtools:::cran_packages()
    deps <- tools::dependsOnPkgs(pkg, dependencies, recursive,
      installed = all)
    deps <- setdiff(deps, ignore)
    devtools:::sort_ci(deps)
  }

## Get the data and format it
all <- packages_all()
deps <-
  revdep_all("SpatialExperiment",
    recursive = TRUE,
    bioconductor = TRUE)
deps_info <- subset(all, Package %in% deps)

## Some entries have break lines
remove_enters <- function(x) {
  gsub("\n", "", x)
}

## Put together in a table
df <- data.frame(do.call(cbind,
  lapply(deps_info[, c("Package", "Title", "URL")], remove_enters)
))
df$URL <- sapply(df$URL, function(x) {
  if(is.na(x)) {
      return(x)
  } else {
      if (!grepl("^http", x)) x <- paste0("https://", x)
      return(paste0("[", basename(x), "](", x, ")"))
  }
})

## Make a little table
knitr::kable(
  df,
  row.names = FALSE,
  caption = "Packages related to `SpatialExperiment`.",
  label = NA
)
```

This table is generated automatically based on all the reverse package dependencies for `SpatialExperiment` @Righelli2021. 

# Contributors

[Lukas M. Weber](https://lmweber.org/), *Johns Hopkins Bloomberg School of Public Health, Baltimore, MD, USA*

[Abby Spangler](https://twitter.com/abspangler), *Lieber Institute for Brain Development, Baltimore, MD, USA*

[Leonardo Collado-Torres](https://twitter.com/lcolladotor), *Lieber Institute for Brain Development, Baltimore, MD, USA*

[Stephanie C. Hicks](https://www.stephaniehicks.com/), *Johns Hopkins Bloomberg School of Public Health, Baltimore, MD, USA*

As well as [GitHub contributors](https://github.com/lmweber/OSTA-book/graphs/contributors) who have helped us improve this book over time. Thank you for your contributions!

# Software

This book would not be possible without an immense list of software and methods contributed by the open source R and Bioconductor communities. 

```{r OSTA_dependencies}
options(width = 120)
pkgs <- installed.packages()[, "Package"]
sessioninfo::session_info(pkgs, include_base = TRUE)
```

The GitHub Actions workflow used in this book is based on [`biocthis`](https://bioconductor.org/packages/biocthis) and is made possible thanks to the [Bioconductor docker images](http://bioconductor.org/help/docker/).