# Dimensionality reduction {#dimensionality_reduction}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble(cache = TRUE)
```


## Overview

Chapter on dimensionality reduction.

```{r, echo=FALSE, results="asis"}
extractCached("human_DLPFC.Rmd", "feature_selection_HVGs", c("spe", "top_hvgs"))
```



## PCA

Apply principal component analysis (PCA) to the set of top highly variable genes (HVGs) to reduce the dimensionality of the dataset, and retain the top 50 principal components (PCs) for further downstream analyses.

This is done for two reasons: (i) to reduce noise due to random variation in expression of biologically uninteresting genes, which are assumed to have expression patterns that are independent of each other, and (ii) to improve computational efficiency during downstream analyses.

We use the computationally efficient implementation of PCA provided in the `scater` package [@McCarthy2017-zd]. This implementation uses randomization, and therefore requires a random seed for reproducibility.

```{r calculate_PCA, message=FALSE}
library(scater)

# calculate PCA
set.seed(123)
spe <- runPCA(spe, subset_row = top_hvgs)

reducedDimNames(spe)
dim(reducedDim(spe, "PCA"))
```



## UMAP

We also calculate UMAP [@McInnes2018-lx] on the set of top 50 PCs and retain the top 2 UMAP components, which will be used for visualization purposes.

```{r calculate_UMAP}
# calculate UMAP on top 50 PCs
set.seed(123)
spe <- runUMAP(spe, dimred = "PCA")

reducedDimNames(spe)
dim(reducedDim(spe, "UMAP"))

# update column names (for easier plotting)
colnames(reducedDim(spe, "UMAP")) <- paste0("UMAP", 1:2)
```



## Plots

Display example plots in PCA and UMAP space. These plots will become more informative in the next section, where we also have cluster labels.

```{r plot_PCA, fig.height=4, fig.width=4.5}
# display PCA plot
plotReducedDim(spe, "PCA")
```

```{r plot_UMAP, fig.height=4, fig.width=4.5}
# display UMAP plot
plotReducedDim(spe, "UMAP")
```


