# Dimensionality reduction


## Background

Chapter on dimensionality reduction



## Previous steps

*Code to run steps from the previous chapters, to generate the `SpatialExperiment` object required for this chapter.*

```{r previous_steps, message=FALSE, class.source = 'fold-hide', results='hide'}
# ---------
# load data
# ---------

library(SpatialExperiment)
library(STexampleData)

spe <- load_data("Visium_humanDLPFC")

# --------------------
# quality control (QC)
# --------------------

library(scater)

# subset to keep only spots over tissue
spe <- spe[, spatialData(spe)$in_tissue == 1]

# identify mitochondrial genes
is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)

# calculate per-spot QC metrics and store in colData
spe <- addPerCellQC(spe, subsets = list(mito = is_mito))

# select QC thresholds
qc_lib_size <- colData(spe)$sum < 500
qc_detected <- colData(spe)$detected < 250
qc_mito <- colData(spe)$subsets_mito_percent > 30
qc_cell_count <- colData(spe)$cell_count > 12

# combined set of discarded spots
discard <- qc_lib_size | qc_detected | qc_mito | qc_cell_count
colData(spe)$discard <- discard

# filter low-quality spots
spe <- spe[, !colData(spe)$discard]

# -------------
# normalization
# -------------

library(scran)

# quick clustering for pool-based size factors
set.seed(123)
qclus <- quickCluster(spe)

# calculate size factors and store in object
spe <- computeSumFactors(spe, cluster = qclus)

# calculate logcounts (log-transformed normalized counts) and store in object
spe <- logNormCounts(spe)

# -----------------
# feature selection
# -----------------

# remove mitochondrial genes
spe <- spe[!is_mito, ]

# fit mean-variance relationship
dec <- modelGeneVar(spe)

# select top HVGs
top_hvgs <- getTopHVGs(dec, prop = 0.1)
```



## Principal component analysis (PCA)

Apply principal component analysis (PCA) to the set of top highly variable genes (HVGs) to reduce the dimensionality of the dataset, and retain the top 50 principal components (PCs) for further downstream analyses.

This is done for two reasons: (i) to reduce noise due to random variation in expression of biologically uninteresting genes, which are assumed to have expression patterns that are independent of each other, and (ii) to improve computational efficiency during downstream analyses.

We use the computationally efficient implementation of PCA provided in the `scater` package [@McCarthy2017]. This implementation uses randomization, and therefore requires setting a random seed for reproducibility.

```{r calculate_PCA}
# compute PCA
set.seed(123)
spe <- runPCA(spe, subset_row = top_hvgs)

reducedDimNames(spe)
dim(reducedDim(spe, "PCA"))
```



## Uniform Manifold Approximation and Projection (UMAP)

We also run UMAP [@McInnes2018] on the set of top 50 PCs and retain the top 2 UMAP components, which will be used for visualization purposes.

```{r calculate_UMAP}
# compute UMAP on top 50 PCs
set.seed(123)
spe <- runUMAP(spe, dimred = "PCA")

reducedDimNames(spe)
dim(reducedDim(spe, "UMAP"))

# update column names for plotting functions
colnames(reducedDim(spe, "UMAP")) <- paste0("UMAP", 1:2)
```



## Visualizations

Generate plots using plotting functions from the [ggspavis](https://github.com/lmweber/ggpavis) package. In the next chapter on clustering, we will add cluster labels to these reduced dimension plots.

```{r reduced_dim_plots, message=FALSE, fig.width=3.75, fig.height=3.75}
library(ggspavis)

# plot top 2 PCA dimensions
plotDimRed(spe, type = "PCA")

# plot top 2 UMAP dimensions
plotDimRed(spe, type = "UMAP")
```


