# (PART) Downstream analyses {-}

# Downstream analyses {#downstream_analyses}

This part contains chapters describing each of the main steps in a downstream analysis pipeline for spatial transcriptomics data. The steps include loading the data, quality control, normalization, feature selection, dimensionality reduction, clustering, and interpretation. The next part ([Extended topics](#extended_topics)) contains additional steps that may be relevant for certain experiments. In [Workflows](#workflows), we also provide condensed versions of the analysis pipeline for several datasets, focusing on relatively standard methods only, and with minimal documentation, to make it easier to see how the steps fit together into a full pipeline.

Throughout the analysis pipeline, we follow the Bioconductor principle of modularity -- the pipeline uses a consistent object structure for the input and output from each step, so if you have an additional alternative method that you wish to try for one of the steps (or are developing one!), then it should be relatively simple to substitute this method and continue with the rest of the pipeline.





## Load data


## Overview

In this chapter, we show how to load and inspect the example dataset used in the rest of the analysis pipeline chapters.

The example dataset is a subset of a 10x Genomics Visium dataset from healthy human brain samples from the dorsolateral prefrontal cortex (DLPFC) region, from our publication @Maynard2020-ke.

For more details on the dataset, see the [Human DLPFC workflow](#human_DLPFC_dataset) chapter.



## Install packages

Install required packages (if not already installed).

```{r unref-install_packages, eval=FALSE}
install.packages("remotes")
remotes::install_github("drighelli/SpatialExperiment")

install.packages("BiocManager")
BiocManager::install("ExperimentHub")
remotes::install_github("lmweber/STexampleData")
```



## Load data

Load the human DLPFC example dataset from the [STexampleData](https://github.com/lmweber/STexampleData) package. Note that this is a subset containing only one sample (sample 151673) from the full dataset. See [Human DLPFC workflow](#human_DLPFC_dataset) for details.

```{r load_data, message=FALSE}
library(ExperimentHub)
library(STexampleData)

ehub <- ExperimentHub()

spe <- load_data("Visium_humanDLPFC")
```



## Inspect object

The example dataset is currently provided as a `SingleCellExperiment` object. We can inspect this object as follows.

[TO DO: update to use `SpatialExperiment`, which also includes the `spatialCoords` slot.]

We see that the object contains 33,538 rows (genes), 3,639 columns (spots or spatial barcodes), one `assay` (`counts` containing raw counts), `rowData` contains columns describing the genes, and `colData` contains columns describing the spots (including `x_coord` and `y_coord` containing the x and y coordinates from the tissue slide).

```{r inspect_object, message=FALSE}
spe
class(spe)
dim(spe)
names(assays(spe))
head(rowData(spe), 3)
head(colData(spe), 3)
```

