# Marker genes


## Background

Chapter on identifying marker genes for clusters by differential gene expression testing.



## Previous steps

*Code to run steps from the previous chapters, to generate the `SpatialExperiment` object required for this chapter. For more details on each step, see the previous chapters.*

```{r previous_steps, message=FALSE, results='hide'}
# LOAD DATA

library(SpatialExperiment)
library(STexampleData)
spe <- load_data("Visium_humanDLPFC")

# QUALITY CONTROL (QC)

library(scater)
# subset to keep only spots over tissue
spe <- spe[, spatialData(spe)$in_tissue == 1]
# identify mitochondrial genes
is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
# calculate per-spot QC metrics
spe <- addPerCellQC(spe, subsets = list(mito = is_mito))
# select QC thresholds
qc_lib_size <- colData(spe)$sum < 500
qc_detected <- colData(spe)$detected < 250
qc_mito <- colData(spe)$subsets_mito_percent > 30
qc_cell_count <- colData(spe)$cell_count > 12
# combined set of discarded spots
discard <- qc_lib_size | qc_detected | qc_mito | qc_cell_count
colData(spe)$discard <- discard
# filter low-quality spots
spe <- spe[, !colData(spe)$discard]

# NORMALIZATION

library(scran)
# quick clustering for pool-based size factors
set.seed(123)
qclus <- quickCluster(spe)
# calculate size factors
spe <- computeSumFactors(spe, cluster = qclus)
# calculate logcounts (log-transformed normalized counts)
spe <- logNormCounts(spe)

# FEATURE SELECTION

# remove mitochondrial genes
spe <- spe[!is_mito, ]
# fit mean-variance relationship
dec <- modelGeneVar(spe)
# select top HVGs
top_hvgs <- getTopHVGs(dec, prop = 0.1)

# DIMENSIONALITY REDUCTION

# compute PCA
set.seed(123)
spe <- runPCA(spe, subset_row = top_hvgs)
# compute UMAP on top 50 PCs
set.seed(123)
spe <- runUMAP(spe, dimred = "PCA")
# update column names
colnames(reducedDim(spe, "UMAP")) <- paste0("UMAP", 1:2)

# CLUSTERING

# graph-based clustering
set.seed(123)
k <- 10
g <- buildSNNGraph(spe, k = k, use.dimred = "PCA")
g_walk <- igraph::cluster_walktrap(g)
clus <- g_walk$membership
colLabels(spe) <- factor(clus)
```



## Marker genes

Identify marker genes by testing for differential gene expression between clusters.

We use the `findMarkers` implementation in `scran` [@Lun2016], using a binomial test, which tests for genes that differ in the proportion expressed vs. not expressed between clusters. This is a more stringent test than the default t-tests, and tends to select genes that are easier to interpret and validate experimentally.


```{r marker_genes, message=FALSE}
library(scran)

# set gene names as row names for visualization purposes
rownames(spe) <- rowData(spe)$gene_name

# test for marker genes
markers <- findMarkers(spe, test = "binom", direction = "up")

# returns a list with one DataFrame per cluster
markers
```


```{r marker_genes_heatmap, message=FALSE, fig.width=5, fig.height=6}
library(pheatmap)

# plot log-fold changes for one cluster over all other clusters
# selecting cluster 1
interesting <- markers[[1]]
best_set <- interesting[interesting$Top <= 5, ]
logFCs <- getMarkerEffects(best_set)

pheatmap(logFCs, breaks = seq(-5, 5, length.out = 101))
```


```{r marker_genes_expression, message=FALSE, fig.width=7, fig.height=7}
library(scater)

# plot log-transformed normalized expression of top genes for one cluster
top_genes <- head(rownames(interesting))

plotExpression(spe, x = "label", features = top_genes)
```


