# Mouse coronal {#mouse_coronal}

```{r unref-setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble(cache = TRUE)
```


This workflow analyzes a mouse coronal brain section dataset measured using the Visium platform. This dataset was generated by 10x Genomics, and the raw data files are publicly available from the [10x Genomics website](https://support.10xgenomics.com/spatial-gene-expression/datasets/1.1.0/V1_Adult_Mouse_Brain).

The workflow follows the same steps as in the main chapters, with adaptations at several points due to characteristics of this dataset.


## Description of dataset {#mouse_coronal_dataset}

This dataset measures transcriptome-wide gene expression on a Visium slide spanning one hemisphere of a mouse coronal brain section. For experimental details, see the [10x Genomics website](https://support.10xgenomics.com/spatial-gene-expression/datasets/1.1.0/V1_Adult_Mouse_Brain).

Due to the small size of the mouse brain and the dimensions of the Visium slide (6.5mm x 6.5mm), the measurements span an entire brain hemisphere. Therefore, we can use this dataset to compare gene expression profiles between major anatomical regions of the mouse brain. In particular, we can can investigate spatial gradients of gene expression within major anatomical regions or cell types. (However, note that due to the resolution of the spots, each anatomical region is represented by only a relatively small number of spots, and it is not possible to resolve fine-grained anatomical features such as cortical layers.)

Since cells in the mouse brain are relatively small, we also expect relatively higher numbers of cells per spot.


## Load data

Load the dataset from the [STdata](https://github.com/lmweber/STdata) package. The data object is formatted in the `VisiumExperiment` object class defined in the [SpatialExperiment](https://bioconductor.org/packages/SpatialExperiment) package.

```{r load_data, message=FALSE}
library(SpatialExperiment)
library(STdata)

# load object
spe <- load_data("mouse_coronal")
spe
```


## Plot data

As an initial check, plot the spatial coordinates (spots) in x-y space.

```{r plot_data, message=FALSE, fig.width=3.5, fig.height=3.5}
library(spatzli)
library(ggplot2)

# plot spatial coordinates (spots)
plotSpots(spe, cluster_id = NULL)
```


