# Normalization {#normalization}


## Overview

Chapter on normalization.

```{r, echo=FALSE, results="asis", cache=TRUE}
# add steps to generate object up to here and cache
```



## Logcounts

Calculate log-transformed normalized counts, using pool-based size factors and deconvolution to spot level.

We use normalization methods for scRNA-seq from `scater` [@McCarthy2017-zd] and `scran` [@Lun2016-dn], treating spots as equivalent to cells. Note that since we have only a single sample, there are no blocking factors in the experimental design.

```{r normalization, fig.height=4, message=FALSE}
library(scater)
library(scran)

# clustering for pool-based size factors
set.seed(123)
clusters <- quickCluster(spe)
table(clusters)

# calculate size factors and store in SCE object
spe <- computeSumFactors(spe, cluster = clusters)

summary(sizeFactors(spe))
hist(sizeFactors(spe), breaks = 20)

# calculate logcounts (log-transformed normalized counts) and store in SCE object
spe <- logNormCounts(spe)

assayNames(spe)
```


